Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-19 paper/paper.tex
Rewriting section 4

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:category-theory}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on
inductive data types.  We might have as well defined them on the subtype of
finite elements of coinductive data types using a predicate which states when an
element of a coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}.
We show this by
defining a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
recursive recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}

\noindent
The main advantage of this definition is that recursive anamorphisms
compose with catamorphisms without the need to reason about termination
and finiteness of values of coinductive types.
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

Suppose that we have an algebra \coq{a : Alg F B}, a recursive coalgebra
\coq{c : RCoalg F A}, and a property \coq{P : A -> B -> Prop}, and we want to
guarantee that the hylomorphism satisfies this property:
\coq{forall x, P x (hylo a c x)}.
Such proofs in our framework are done by induction on the proof that the
recursive coalgebra terminates. In particular, the induction hypothesis would be
that \coq{P} is satisfied by any recursive call of the hylomorphism
\begin{coqcode}
forall e, P (cont (c x) e) (hylo a c (cont (c x) e))
\end{coqcode}
\noindent
Then, by the uniqueness property of hylomorphisms, it is enough to prove that
\begin{coqcode}
P x ((a \o fmap (hylo a c) \o c) x)
\end{coqcode}
We will see in Section~\ref{sec:examples} one example where we will prove that
the output of a sorting algorithm is a sorted permutation of the input, and we
show that these proofs are comparable to other techniques for
non-structural recursion.

@@

\subsection{Sorting Algorithms}\label{sec:sorting}
Our first case study is divide-and-conquer sorting algorithms.  Encoding them
in our framework will require the use of recursive hylomorphisms and
termination proofs.  We complete the sorting algorithm examples by applying
fusion optimisation.

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is
split into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
These define a container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
% \mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}
The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with
the following algebra and coalgebra.
\begin{coqcode}
  Definition merge : App (TreeF unit int) (list int) ~> list int.
|{ x : (App (TreeF unit int) (list int)) ~> (
           match x with
           | MkCont sx kx =>
               match sx return (Container.Pos sx -> _) -> _ with
               | Leaf _ _ => fun _ => nil
               | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
               end kx
           end
  )}|.
Defined.
Definition split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge split} is the following:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.
The mergesort algorithm can be defined analogously.

@@

We show that the output of \coq{hylo merge split} produces a sorted
permutation of the input:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge split l) /\ Perm l (hylo merge split l).
\end{coqcode}
The proof in our framework starts by induction on the proof that \coq{split}
is recursive, and that it terminates on \coq{l}, i.e. using tactic
\coq{induction (recP split l)}. Then, we can unfold
\coq{hylo merge split} to \coq{merge \o fmap (hylo merge split) \o split} using
hylomorphism uniqueness, and the proof would then proceed equally to any other
similar proof. Overall, our proof is shorter than alternative implementations, and of comparable complexity\footnote{The QSort definition and
correctness proof in the equations package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
As an example of how can we use program calculation techniques in our 
framework, we show how another traversal can be fused into the
divide-and-conquer algorithm using the laws of hylomorphisms.
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge split.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is \coq{hylo (merge \o natural times_two) split}. In this definition,
\coq{natural} defines a natural transformation by applying \coq{times_two} to
the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
two. Our formalisation contains a proof that \coq{natural} is indeed a natural
transformation, which relies on the fact that it preserves the structure of the
shapes and, therefore, the validity of the positions.  The extracted OCaml code
is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

>>>>

\subsubsection{On the subtype of finite elements}
As we mention in
Section~\ref{sec:category-theory}, we define recursive anamorphisms as
hylomorphisms built by using a recursive coalgebra, and the respective
initial \coq{F}-algebra.
In other words, in this development we have defined recursive anamorphisms on
inductive data types.  We might have as well defined them on the subtype of
finite elements of coinductive data types using a predicate which states when an
element of a coinductive data type is finite:
\begin{coqcode}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{coqcode}
Now the subtype \coq{{x : GFix F | FinF x}} of finite elements for \coq{GFix F}
is isomorphic its corresponding the inductive data type \coq{LFix F}.
We show this by
defining a catamorphism \coq{ccata_f_} from the
subtype \coq{{x : GFix F | FinF x}} of finitary elements of \coq{GFix F} to
any \coq{F}-algebra.
\begin{coqcode}
Definition ccata_f_ `{eA : setoid A} (g : Alg F A)
  : forall x : GFix F, FinF x -> A := fix f x H :=
    let hx := g_out x in
      g (MkCont (shape hx) (fun e => f (cont hx e) (FinF_inv H e))).
\end{coqcode}
We now prove this is isomorphic to the least fixed-point of the functor
\coq{F}.
We take the catamorphism from the finite elements of \coq{GFix F} to the
inductive data type \coq{LFix F} using the \coq{F}-algebra \coq{l_in}. Its
inverse is the catamorphism on the restriction of \coq{g_in} to the finite
elements of \coq{GFix}, which we denote by \coq{lg_in}.  The following lemmas
prove the isomorphism:
\begin{coqcode}
Lemma cata_ccata `{setoid A} : cata lg_in \o ccata l_in =e id.
Lemma ccata_cata `{setoid A} : ccata l_in \o cata lg_in =e id.
\end{coqcode}
The finite subtype of \coq{GFix F} allows us to compose
catamorphisms and anamorphisms, by using the above isomorphism. In our work,
however, we use \emph{recursive} anamorphisms, defined as hylomorphisms on the
recursive recursive coalgebra \coq{c}, and the initial \coq{F}-algebra:
\begin{coqcode}
Definition rana c := hylo l_in c.
\end{coqcode}
The main advantage of this definition is that recursive anamorphisms compose
with catamorphisms without the need to reason about termination and finiteness
of values.
% This predicate is equivalent to the proposition that
% the anamorphism of such coalgebras produce only finite trees:
% \begin{coqcode}
% Lemma ana_rec_term `{setoid A} (h : Coalg F A)
%   : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
% \end{coqcode}

@@

Suppose that we have an algebra \coq{a : Alg F B}, a recursive coalgebra
\coq{c : RCoalg F A}, and a property \coq{P : A -> B -> Prop}, and we want to
guarantee that the hylomorphism satisfies this property:
\coq{forall x, P x (hylo a c x)}.
Such proofs in our framework are done by induction on the proof that the
recursive coalgebra terminates. In particular, the induction hypothesis would be
that \coq{P} is satisfied by any recursive call of the hylomorphism
\begin{coqcode}
forall e, P (cont (c x) e) (hylo a c (cont (c x) e))
\end{coqcode}
\noindent
Then, by the uniqueness property of hylomorphisms, it is enough to prove that
\coq{P} is satisfied by the unfolding of the hylomorphism:
\begin{coqcode}
P x ((a \o fmap (hylo a c) \o c) x)
\end{coqcode}
We will see in Section~\ref{sec:examples} one example where we will prove that
the output of a sorting algorithm is a sorted permutation of the input, and
show that these proofs are comparable to other techniques for non-structural
recursion, with the additional advantage that any result that is proven on a
hylomorphism will also hold for any result of doing program calculation.

@@

\subsection{Sorting Algorithms}\label{sec:sorting}
Our first case study is divide-and-conquer sorting algorithms encoded as
recursive hylomorphisms. Our accompanying code contains both mergesort and
quicksort as examples, but we will focus on quicksort for conciseness. We will
also show how to use program calculation to apply a fusion optimisation.  The
recursive structure of both mergesort and quicksort is a binary tree. For
example, in the case of quicksort, a list is split into a pivot, the label of
the node, and two sublists. We define the data functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
To define an equivalent container, we notice that this functor has two shapes,
one for leaves, and one for nodes, and that nodes have two positions (akin to
\coq{l r : X} in \coq{i_node}). We also note that, in this container, positions
are only valid in nodes.
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We name \coq{TreeF} the container that has these shapes and (base) poisitions.
For convenience, we define a series of wrappers to construct these containers:
% \mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Notation a_leaf x := (MkCont (Leaf _ x) (@dom_leaf _ _ _ x)).
Notation a_node x l r := (MkCont (Node _ x)
  (fun p => match val p with | Lbranch => l | Rbranch => r end)).
\end{coqcode}
The container for the Quicksort hylomorphism is \coq{TreeF unit int}, with the
following algebra and coalgebra. In the below definition, we use \coq{posL} and
\coq{posR} as wrappers for \coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition merge_f (x : App (TreeF unit int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
Definition qsplit_f (x : list int) : App (TreeF unit int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.
\end{coqcode}
Our tactics automatically discharge the proof that these are proper morphisms
(called \coq{merge} and \coq{qsplit}), and we prove that the coalgebra
\coq{qsplit} is recursive by showing that it respects the ``less-than'' relation
on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge qsplit} is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}

@@

We show that the output of \coq{hylo merge qsplit} produces a sorted
permutation of the input:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
The proof in our framework starts by induction on the proof that \coq{qsplit}
is recursive, and that it terminates on \coq{l}, i.e. using tactic
\coq{induction (recP qsplit l)}. Then, we can unfold
\coq{hylo merge qsplit} to \coq{merge \o fmap (hylo merge qsplit) \o qsplit} using
hylomorphism uniqueness, and the rest of the proof is standard. Overall, our
proof is shorter than alternative implementations, and of comparable
complexity\footnote{The QSort definition and correctness proof in the equations
package is under 200LOC --
\href{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v}{https://github.com/mattam82/Coq-Equations/blob/8.20/examples/quicksort.v},
and other online proofs of correctness are of similar complexity
\href{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}{https://gist.github.com/RyanGlScott/ff36cd6f6479b33becca83379a36ce49}}.
% and can be found in the
% formalisation\footnote{https://github.com/dcastrop/coq-hylomorphisms}. % [REF LINK].

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
As an example of how can we use program calculation techniques in our 
framework, we show how another traversal can be fused into the
divide-and-conquer algorithm using the laws of hylomorphisms.
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge qsplit.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting and \coq{hylo_fusion_l} to fuse
\coq{times_two} into \coq{hylo merge qsplit}.
After the necessary rewrites, we obtain \coq{hylo (merge \o natural times_two) qsplit}. In this definition,
\coq{natural} defines a natural transformation by applying \coq{times_two} to
the shapes, thus multiplying every pivot by
two.
%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
The extracted OCaml code
is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
